Deep learning on psychological arousal classification

### 2015-05-07 paper/paper.tex
Techniques

\subsection{EEG}
The EEG (electroencephalogram) signal is a voltage signal that can be measured on the surface of the scalp, arising from large areas of coordinated neural activity. Rhythmic fluctuations (oscillation patterns) in the EEG signal occur within several particular frequency bands, and the relative level of activity within each frequency band has been associated with brain/emotional states such as focused attentional processing, engagement, and frustration (Marosi et al., 2002; Marosi et al., 2001;
Mundy-Castle, 1951; Berka et al., 2007) and cognitive and memory performance (Gruber et al., 2004; Fernandez et al., 1998; Gevins et al., 1997; Jensen et al., 2007; Klimesch, 1999). The discriminating power of EEG has been demonstrated in a number of studies, achieving 86\% accuracy on 14-way task discrimination, 95\% accuracy on discriminating high/low memory load (Gevins et al., 1998). 
\subsection{IAPS}
\subsection{Deep Learning}
Contemporary deep learning was introduced relatively recently (Hinton, 2006). Prior to the advent of deep learning, classification algorithms by in large operated on some pre-determined preprocessing steps which transforms the raw input into features. These features were then used by a learning algorithm to train a classifier. Deep algorithms, by contrast, combine data processing and classification in the same learning pipeline. This allows the system to learn how to process raw data from
patterns in the data rather than relying on hand-engineered pre-processing steps. This is why deep learning has alternatively been called feature learning or representation learning.
Deep learning excels when it is trained on large repositories of data. So the relatively recent availability of large datasets for speech, computer vision, and text processing has allowed deep learning researchers to train models matching or beating state-of-the-art systems (Boulanger-Lewandowski et al., 2012; Krizhevsky et al., 2012; Miklov et al., 2011).  Deep learning is particularly good at extracting structure and information from raw data with little to no knowledge about the
nature of the data or the nature of the intended classification task by performing unsupervised pre-training (Mesnil et al., 2012). Researchers hypothesize that this is because deep learning algorithms are able to disentangle the many sources of variation (Bengio, 2009).  This is supported by empirical analysis of the learnt features which demonstrate that features that best distinguish task domains are largely separate from features that best distinguish classification conditions (Glorot et
al., 2011), a separation that suggests features are selectively sensitive to only one type of variation.  The usefulness of this disentanglement is evident in the strong performance of deep learning algorithms in transfer learning competitions (Mesnil et al .,2012; Goodfellow et al., 2011).
Disentangling sources of variation is a particularly useful trait for use in EEG. As mentioned earlier, sources of signal in EEG are not well-understood and feature engineering is very difficult. By providing deep learning algorithms with a sufficiently large EEG dataset, the algorithms may be able to disentangle the many sources of noise and signal to recognize the specific patterns associated with mental states of interest and discarding patterns associated with environmental noise.
Further, deep algorithms excel at transferring between different task domains which may be helpful in creating classifiers that can be used on unseen subjects despite non-trivial differences in subject psychology and physiology.
\subsection{Deep Learning and EEG}
Deep learning has also had some very recent success in non-EEG biometrics and laboratory EEG. Mirowski et al. (2008) used a convolutional neural network on intracranial EEG signal, predicting seizures with zero-false-alarm rate on 20 out of 21 patients, beating SVM and logistic regression. Cecotti and Gräser (2010) achieved state-of-the-art results on detecting the P300 event-related-potential with laboratory grade EEG also using convolutional networks. And more recently, Martinez et al.
(2013) used deep learning techniques to produce models of affect from non-EEG biometrics (Skin Conductance and Blood Volume Pulse) which perform significantly better than ad-hoc feature models.
Uniquely, we are using Deep Learning techniques on non-medical-grade EEG devices to track emotion.

\section{Data} \label{section:data}
Our experiment uses data from Siegle et al. (forthcoming). To our knowledge, this is the largest dataset of EEG collected in controlled laboratory conditions (TODO: get x and n) with n unique subjects from x different sites. The dataset contains various stimuli types and our experiments focus on the IAPS stimuli (TODO: cite lang et al 2008). IAPS is a set of images designed to illicit emotional responses along two emotional axes - high, low arousal and positive, negative valence. The stimuli
fall into 4 categories - low arousal positive valence, low arousal negative valence, high arousal positive valence, and high arousal negative valence. In this dataset, each subject viewed 10 images from each category with a 2 (TODO: check this) second fixation period between each stimulus category. One limitation of the experiment was that the images were shown to the subject in the same sequence.

The data was processed with %TODO: asdfasdfasdf how did siegle process the data??
%TODO: we have big data, oooh. okay maybe not that big but it still takes forever to load 
%-Dataset (details)
% -processing
%-Features
% -raw
% -dtft
%  -windows
% -wavelets
% -other

\section{Techniques} \label{section:techniques}
%TODO: we thought technology x would be a great idea because it has these advantages over the sucky way that one guy was doing it. so we wrote some code and did it. we wrote the code. :o
%-Deep learning
% -deep nets
% -dropout
% -autoencoder
% -denoising autoencoder
% -convolutional neural nets
% -parameterized rectifiers
Parameterized rectified linear units (PReLUs) \cite{DBLP:journals/corr/HeZR015} are leaky ReLUs whose slopes are learnable parameters.
% -batchnorm (plus all that other stuff)
Batch normalization \cite{DBLP:journals/corr/IoffeS15} is a method for training deep neural networks that makes normalization a part of the network, allowing for better performance and training speed.
%-Ensemble
% -random forest
% -classifier fusion

>>>>

%TODO: so like our thing has never actually been done before. actually, this guy did this one thing but it really sucks. so we were like obviously we could do better with technology x.
%-EEG
% -what is EEG
% -EEG patterns (psych stuff/neuro stuff)
% -Dataset (brief)
%-Deep learning
% -what is deep learning
% -deep learning + EEG

\section{Data} \label{section:data}
%TODO: we have big data, oooh. okay maybe not that big but it still takes forever to load 
%-Dataset (details)
% -processing
%-Features
% -raw
% -dtft
%  -windows
% -wavelets
% -other

\section{Techniques} \label{section:techniques}
\subsection{Deep nets}
Deep nets, neural nets composed of many layers, are known to be able to represent highly complex decision boundaries when used with non-linear activation functions.

In our models, we use the rectified linear unit (ReLU) \cite{ICML:2010NairH10}
\[ \mathrm{ReLU}(x) = \max(0, x) \]
as the activation function for the hidden layers. This non-linear function comes with the advantage that it and its derivatives are quick to compute. However, it is not differentiable at zero; this introduces no effects in practice.

One may use a ``leaky'' ReLU
\[ \mathrm{ReLU}(x; k) = \max(kx, x) \]
where the negative arm of the function has a slightly positive slope $0 < k \ll 1$, ensuring non-zero derivatives.

Parameterized rectified linear units (PReLUs) \cite{DBLP:journals/corr/HeZR015} are leaky ReLUs whose slopes are learnable parameters.

Other standard candidates are the sigmoid function
\[ \sigma(x) = \frac{1}{1+e^{-x}} \]
and the hyperbolic tangent function
\[ \tanh(x) = \frac{e^x - e^{-x}}{e^x + e^{-x}} \]
as they are differentiable everywhere.

% TODO: describe the layers/architecture more in detail

\subsection{Autoencoders}
Autoencoders are a class of neural network architectures employed for use in representation learning. Autoencoders are fully connected neural networks containing at least one hidden layer, an encoder, and a decoder. The transformation described by the weights going from the input units to the middle hidden layer is the encoder, and the transformation described by the weights going from the middle hidden layer to the output units is the decoder. Typically, the decoder weights are shared as the transpose of the encoder weights with a separate set of bias units.

Autoencoders can be trained in an unsupervised manner by learning the identity function on a set of data. An autoencoder with multiple hidden layers can be represented as a composition of multiple autoencoders, each with one hidden layer. These stacked autoencoders can be trained in a layerwise fashion, i.e. by greedily training the outermost autoencoder, then combining with the next inner autoencoder, initializing with the pretrained weights. So, we consider the behavior of autoencoders with one hidden layer: depending on the dimension of the hidden layers compared to the dimension of the input and output layers, autoencoders fall into two realms of behaviors.

\subsubsection{Hidden layer smaller than input layer}
If the hidden layer has a smaller dimension than the input layer, then in order to accurately approximate the identity function, the autoencoder learns to compress the larger input space into the smaller vector space represented by the hidden layer activations and retain salient features. If linear activation units $f(x) = x$ are used, then autoencoding is equivalent to principal component analysis (PCA). The trained encoder weights can be used to initialize a subset of a larger neural net classifier to expose the same learned features.

\subsubsection{Hidden layer larger than input layer}
If the hidden layer has a larger dimension than the input layer, straightforward training does not learn a meaningful representation because the identity encoding is a trivial optimum for training and will overfit. To force a useful representation, we introduce noise in the input and try to get the network to learn the original output. This causes the network to produce distributed representations of the input data that are useful in reconstructing the data. By adding Gaussian distributed noise to the input vectors during preprocessing, we obtain a denoising autoencoder that becomes resistant to noise. Another method of adding noise is dropout: during training, a fraction $p$ of the input vector dimensions (usually $p = 0.5$) is randomly set to $0$. During testing, the entire input vector is used and the layer weights are multiplied by $\frac{1}{p}$ to restore scaling of activations.

\subsection{Training and batch normalization}
Training is typically done on batches of inputs at once to take advantage of data parallelism techniques to speed up training.
Batch normalization \cite{DBLP:journals/corr/IoffeS15} is a method for training deep neural networks that incorporates normalization a part of the network, allowing for better performance and training speed. In each layer, with each batch that is propagated through the network, the mean and variance is taken within the batch to renormalize each example. Two learnable coefficients $\alpha$ and $\beta$ are trained to rescale the normalized values.
At test time, a global mean and variance is taken over samples drawn from the training set to be used to scale test examples.

\subsection{Ensemble learning}
% TODO
%-Ensemble
% -random forest
% -classifier fusion
